Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-02-16 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
added to the bayesian section

% HS: temporary title. Feel free to add suggestions
% \title{Probing higher-curvature gravity theories from binary black hole ringdown signals}
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

% We will consider several modified gravity theories as applications of the \pSEOB{}
% waveform model presented in Sec.~\ref{sec:review_pSEOB}.
%
We briefly review which theories we consider, what are the currents
observational constraints in each of them and what we known about the QNMs of
black holes in each of them.

\subsection{scalar-Gauss-Bonnet gravity}
%
This theory is given by the following action,
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \mathscr{G}
    \right],
\end{equation}
%
where $R$ is the Ricci scalar associated with the metric $g_{\alpha\beta}$ and
$\varphi$ is a dynamical scalar field which couples to the Gauss-Bonnet
invariant $\mathscr{G}$,
%
\begin{equation} \label{eq:def_gb}
    \mathscr{G} =
    R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
    - 4 R^{\mu\nu}R_{\mu\nu}
    + R^2,
\end{equation}
%
with strength set by the dimensionful coupling constant $\ell_{\rm GB}$ with dimensions
of length.

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their black hole solutions.
%
In the first class, the first derivative of the coupling
function $f'(\varphi) = \dd f  / \dd \varphi$ is always nonzero and black holes
are known to always support (secondary) scalar hair.
%
Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
$f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, one can show that theory admits the same stationary,
asymptotically flat black hole solution of GR and scalarized black
holes~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} couplings.
%
Here will consider only the dilatonic theory.

Black holes in both classes support monopolar scalar hair and thus, when in binaries,
can source scalar dipole radiation and are therefore prone to be constrained with
GWs observations of compact binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
\dots

The calculation of the QNMs in this theory is more complicated than in GR already for
nonrotating black holes. The reason is due to the coupling between scalar field and
the Gauss-Bonnet invariant, which manifests into a coupling between scalar perturbation
and gravitational perturbations of polar parity \dots

@@

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of black holes in scalar-Gauss-Bonnet gravity.}
\end{figure}

@@

The theory admits the Schwarzschild black hole solution of GR, but predicts
that spinning black holes have (secondary) scalar hair, which, unlike in
scalar-Gauss-Bonnet gravity, is dipolar~\cite{Yunes:2009hc,Konno:2009kg}.
%
Consequently, the leading-scalar radiation channel is quadrupolar making this
theory presently unconstrained by from the GWs from the inspiral of black hole
binaries alone~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref~\cite{Silva:2020acr} through a
combination information of x-ray observations of the isolated neutron star PSR
J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by the
NICER~\cite{Riley:2019yda,Miller:2019cac} and gravitational-wave observation of
the binary neutron star event GW170817~\cite{TheLIGOScientific:2017qsa}.

Similarly to the case of scalar-Gauss-Bonnet gravity, the perturbations of nonrotating
black holes in dynamical Chern-Simons gravity also have a coupling between scalar and
metric perturbations. However, the coupling is between the scalar and gravitational perturbations
of axial parity~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb,Wagle:2021tam}.

@@

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of black holes in dynamical Chern-Simons gravity.}
\end{figure}

@@

This theory is given by the following action,
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \footnotesize{\sum}_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right],
\end{equation}
%
where $\ell$ is some length scale, assumed to be small compares to the length
scale associated with a black hole, i.e., $\ell / M \ll 1$, and
$L^{(2n)}$ are corrections to the Einstein-Hilbert term action that
introduce involving higher-order curvature tensors (with $2n$ metric
derivatives).

@@

Black holes in these theories were found in Refs.~\cite{deRham:2020ejn,Cano:2020cao} for the dimension-six EFT
and in Refs.~\cite{Cardoso:2018ptl} for the dimension-eight EFT.

@@

% This action encapsulates various modified gravity theories, which often are studied on their own,
% namely,
% %
% \begin{itemize}
%     \item scalar Gauss-Bonnet gravity (sGB): corresponds to $\alpha_{\rm GB}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item dynamical Chern-Simons (dCS): this corresponds to $\alpha_{\rm CS}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item cubic effective field theory of GR (cEFTofGR): this corresponds to $\lame$, $\lamo$ nonzero
%           and all other dimensionless coupling constants equal to zero.
%     \item quartic effective field theory of GR (qEFTofGR): this corresponds to $\varepsilon_{i}$
%           ($i = 1,2,3$) nonzero and all other dimensionless coupling constants equal to zero.
% \end{itemize}
%
% In each of these theories, slowly-rotating black hole and their quasinormal mode spectra
% to leading order in spin have been calculated.
% %
% We can use this results to establish a mapping between theory-specific calculations and
% the theory-agnostic {\sc Parspec} parametrization.
% %
% We discuss how we do this in the section.
%
% In Eqs.~\eqref{eq:action_mods}, $\alpha_{\rm GB}$, $\alpha_{\rm CS}$, $\lame$,
% $\lamo$ and $\varepsilon_{i}$ are dimensionless constants.

@@

In this formalism our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\})\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\})
\end{align}
\end{subequations}
%
where the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict the final mass and spin $(M_f,\, \chi_f)$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
Additionally, $(\omega_k,\, \tau_k)$ depend on the coupling constant $\ell$ defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$, where $j$ ranges from 0 to the power of spin in Eqs.~\ref{eq:kerr_expansion} upto which we include corrections.
%
For example, if we restrict ourselves to corrections in the $j=0$, or $\chi_f^0$ terms, then we just have two additional degrees of freedom, $(\omega_k^{(0)}, \tau_k^{(0)})$ per mode $k$. We will call this parametrized waveform model $\pSEOB$ in this paper.
%
In this paper, we will restrict ourselves to the $k=(2,2,0)$ QNM and corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$. Assuming a prior which is uniform on the beyond-GR parameters, $(\ell, \{\delta \omega_k^{(j)}\},\{\delta \tau_k^{(j)}\})$, we employ a Bayesian framework to stochastically sample over the parameter space using the \texttt{LALInference} algorithm.~\cite{lallsuite}.
%
The bounds on these beyond-GR parameters, for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}, are provided in Sec.~\ref{sec:results}.

\iffalse
In the $\SEOB$ model constructed in Ref.~\cite{Cotesta:2018fcv}, the
complex frequencies $\sigma_{l m 0}$ are expressed in terms of the
final BH mass and spin~\cite{Berti:2005ys,Berti:2009kk}, and the
latter are related to the BBH's component masses and spins through
NR--fitting-formulas obtained in
GR~\cite{Taracchini:2013rva,Hofmann:2016yih}. Here instead, in the
spirit of what was done in \paperone, we promote the QNM (complex)
frequencies to be free parameters of the model, while keeping the
inspiral-plunge modes $h_{l m}^\mathrm{inspiral-plunge}(t)$ fixed
to their GR values. More explicitly, we introduce a parameterized
version of the $\SEOB$ model where the frequency and the
damping time of the ${l m 0}$ mode (i.e, $(\omega_{l m 0}, \tau
_{l m 0})$) are defined through the fractional deviations, $(\delta
\omega_{l m 0},\delta \tau_{l m 0})$, from the corresponding GR
values~\cite{Gossan:2011ha,Meidam:2014jpa}.

Thus,
\begin{subequations}
\begin{eqnarray}
\omega_{l m 0} &=& \omega_{l m 0}^{\text{GR}}\, (1 + \delta \omega_{l m 0})\,,\label{eq:nongr_freqs_a} \\
\tau _{l m 0} &=& \tau _{l m 0}^{\text{GR}}\, (1 + \delta \tau_{l m 0})\,. \label{eq:nongr_freqs_b}
\end{eqnarray}
\end{subequations}

The GR quantities $( \omega_{l m 0}^{GR},\tau_{l m 0}^{GR})$ are
constructed using the NR--fitting--formula from Refs.~\cite{Taracchini:2013rva,Hofmann:2016yih}, and are functions of the initial masses and spins, $(m_1, m_2, \chi_1, \chi_2)$. Hence,
\begin{subequations}
\begin{eqnarray}
\omega_{l m 0} &=& \omega_{l m 0}(m_1, m_2, \chi_1, \chi_2, \delta \omega_{l m 0}, \delta \tau_{l m 0})\,,\\
\tau _{l m 0} &=& \tau _{l m 0}(m_1, m_2, \chi_1, \chi_2, \delta \omega_{l m 0}, \delta \tau_{l m 0})\,.
\end{eqnarray}
\end{subequations}
We denote such a parameterized waveform model $\pSEOB$~\footnote{This
waveform model is called {\tt pSEOBNRv4HM} in LAL.}. We note that when leaving $\sigma_{l m}$ to vary
freely, the functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l
  m}(t)$ in general also differ from the GR predictions, since
those functions depend on the QNM complex frequencies, as can be seen
from the expressions for $c_{i,c}^{l m}$ and $d_{1,c}^{l m}$ in Eqs.~(\ref{c1}),
(\ref{c2}), and (\ref{d1}). As a consequence, the ringdown signal (amplitude and phase)
soon after merger deviates from the one predicted by GR.
\fi

@@

\subsection{From theory-agnostic to theory-specific QNM results}

@@

Now as we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating black holes
is modified gravity theories come in two families depending on their transformation upon a parity
transformation: axial and polar. Which is the one we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work on a theory-by-theory case and in each theory perform a
a translation between the metric perturbations $h_{\mu\nu}$ in the Regger-Wheeler-Zerilli gauge and
connect it with the transverse-traceless gauge used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion in Ref.~\cite{Hui:2021cpm} for a counterexample).
and thus how axial and polar gravitational QNMs appear in the GW signal has to be answered in theory-by-theory.
%
This is outside the scope of this paper and here we take the more pragmatic approach
of simply choosing the \emph{least damped gravitational mode between the two parities.}
%
The justification is this is the mode which, if excited during the merger, is
the most likely to appear in the signal.

@@

%%%%%%%%%%%%%%%%%
\section{Analysis using LIGO-Virgo data}
\label{sec:results}
%%%%%%%%%%%%%%%%%

@@

We will consider the events GW150915~\cite{LIGOScientific:2016aoc} and GW200129~\cite{LIGOScientific:2021djp}.
%
They are suitable for ringdown studies due to the high-masses and high signal-to-noise ratios (SNR).

\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $d$ is the data and $\bm{\lambda}$ is a set of vectors that fully
characterize the waveform.
%
$P(\bm{\lambda} \vert \mathcal{H})$ is prior probability distribution,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence.

Our set of parameters is $\bm{\lambda} = \bm{\theta}_{\rm GR} \cup \bm{\theta}_{\rm nGR}$, where
$\bm{\theta}_{\rm nGR} = \{\ell_{i}\}$ is a set of beyond-GR coupling constants.
%
As discussed in Sec.~\ref{sec:review_theories} there is a single coupling
constants in the cases of EdGB and dCS theories, but two in the cubic and
quartic EFTofGR.

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln \mathcal{L}(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk
denotes complex conjugation and $S_{n}(f)$ is the power spectrum density
of the detector.

With the posterior distribution function $P(\bm{\lambda} | d, \mathcal{H})$ determined, we can then
marginalize over all nuisance parameters to obtain the posterior
distribution function the $i$-th beyond-GR coupling constants $\ell_{i}$,
%
\begin{equation}
P(\ell_{i} | d, \mathcal{H}) =
\int \dd \xi_{\rm GR} \,
P(\bm{\lambda} | d, \mathcal{H}) \,.
\label{eq:marg_posterior}
\end{equation}
%

>>>>

% HS: temporary title. Feel free to add suggestions
% \title{Probing higher-curvature gravity theories from binary BH ringdown signals}
\title{Black-hole ringdown as a probe of higher-curvature gravity theories}

@@

% We will consider several modified gravity theories as applications of the \pSEOB{}
% waveform model presented in Sec.~\ref{sec:review_pSEOB}.
%
We briefly review the theories we consider in this paper, what the currents
observational constraints are in each of them and what we know about BH QNMs in each of them.

\subsection{scalar-Gauss-Bonnet gravity}
%
This theory is given by the following action,
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \mathscr{G}
    \right],
\end{equation}
%
where $R$ is the Ricci scalar associated with the metric $g_{\alpha\beta}$ and
%$\varphi$ is a dynamical scalar field which couples to the Gauss-Bonnet
invariant $\mathscr{G}$,
%
\begin{equation} \label{eq:def_gb}
    \mathscr{G} =
    R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
    - 4 R^{\mu\nu}R_{\mu\nu}
    + R^2,
\end{equation}
%
with strength set by the dimensionful coupling constant $\ell_{\rm GB}$ with dimensions
of length. Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH solutions.
%
In the first class, the first derivative of the coupling
function $f'(\varphi) = \dd f  / \dd \varphi$ is always nonzero and BHs
are known to always support (secondary) scalar hair.
%
Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
$f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, one can show that theory admits the same stationary,
asymptotically flat BH solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} couplings.
%
Here we will consider only the dilatonic theory.

BHs in both classes support monopolar scalar hair and thus, when in binaries,
can source scalar dipole radiation and are therefore prone to be constrained with
GWs observations of compact binaries~\cite{Nair:2019iur,Perkins:2021mhb}.
%
\dots

The calculation of the QNMs in this theory is more complicated than in GR already for
nonrotating BHs. The reason is due to the coupling between scalar field and
the Gauss-Bonnet invariant, which manifests into a coupling between scalar perturbation
and gravitational perturbations of polar parity \dots

@@

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of BHs in scalar-Gauss-Bonnet gravity.}
\end{figure}

@@

The theory admits the Schwarzschild BH solution of GR, but predicts
that spinning BHs have (secondary) scalar hair, which, unlike in
scalar-Gauss-Bonnet gravity, is dipolar~\cite{Yunes:2009hc,Konno:2009kg}.
%
Consequently, the leading-scalar radiation channel is quadrupolar making this
theory presently unconstrained by from the GWs from the inspiral of BH
binaries alone~\cite{Nair:2019iur,Perkins:2021mhb}.
%
However, the theory has been constrained by Ref~\cite{Silva:2020acr} through a
combination information of x-ray observations of the isolated neutron star PSR
J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by the
NICER~\cite{Riley:2019yda,Miller:2019cac} and gravitational-wave observation of
the binary neutron star event GW170817~\cite{TheLIGOScientific:2017qsa}.

Similarly to the case of scalar-Gauss-Bonnet gravity, the perturbations of nonrotating
BHs in dynamical Chern-Simons gravity also have a coupling between scalar and
metric perturbations. However, the coupling is between the scalar and gravitational perturbations
of axial parity~\cite{Yunes:2007ss,Cardoso:2009pk,Molina:2010fb,Wagle:2021tam}.

@@

    \node at (+1, 1.25) {${\cal O}(\chi^{1})$};
    \node at (+1, +0.5) {\dots};
    \node at (+1, -0.5) {\dots};
\end{tikzpicture}
\caption{Quasinormal modes of BHs in dynamical Chern-Simons gravity.}
\end{figure}

@@

This theory is given by the following action,
%
\begin{equation} \label{eq:action_eft}
    S_{\rm EFT} = \frac{1}{16 \pi}
    \int \dd^4x \sqrt{-g}
    \left[ R
    +
    \footnotesize{\sum}_{n \geqslant 2} \ell^{2n - 2} L^{(2n)}
    \right],
\end{equation}
%
where $\ell$ is some length scale, assumed to be small compares to the length
scale associated with a BH, i.e., $\ell / M \ll 1$, and
$L^{(2n)}$ are corrections to the Einstein-Hilbert term action that
introduce involving higher-order curvature tensors (with $2n$ metric
derivatives).

@@

BHs in these theories were found in Refs.~\cite{deRham:2020ejn,Cano:2020cao} for the dimension-six EFT
and in Refs.~\cite{Cardoso:2018ptl} for the dimension-eight EFT.

@@

% This action encapsulates various modified gravity theories, which often are studied on their own,
% namely,
% %
% \begin{itemize}
%     \item scalar Gauss-Bonnet gravity (sGB): corresponds to $\alpha_{\rm GB}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item dynamical Chern-Simons (dCS): this corresponds to $\alpha_{\rm CS}$
%           nonzero and all other dimensionless coupling constants equal to zero.
%     \item cubic effective field theory of GR (cEFTofGR): this corresponds to $\lame$, $\lamo$ nonzero
%           and all other dimensionless coupling constants equal to zero.
%     \item quartic effective field theory of GR (qEFTofGR): this corresponds to $\varepsilon_{i}$
%           ($i = 1,2,3$) nonzero and all other dimensionless coupling constants equal to zero.
% \end{itemize}
%
% In each of these theories, slowly-rotating BH and their quasinormal mode spectra
% to leading order in spin have been calculated.
% %
% We can use this results to establish a mapping between theory-specific calculations and
% the theory-agnostic {\sc Parspec} parametrization.
% %
% We discuss how we do this in the section.
%
% In Eqs.~\eqref{eq:action_mods}, $\alpha_{\rm GB}$, $\alpha_{\rm CS}$, $\lame$,
% $\lamo$ and $\varepsilon_{i}$ are dimensionless constants.

@@

In this formalism our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\})\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\})
\end{align}
\end{subequations}
%
where the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict the final mass and spin $(M_f,\, \chi_f)$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
Additionally, $(\omega_k,\, \tau_k)$ depend on the coupling constant $\ell$ defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$, where $j$ ranges from 0 to the power of spin in Eqs.~\ref{eq:kerr_expansion} upto which we include corrections.
%
For example, if we restrict ourselves to corrections in the $j=0$, or $\chi_f^0$, terms for just the $k=(2,2,0)$ mode, then we would just have two additional degrees of freedom, $(\delta \omega_0^{(0)}, \delta \tau_0^{(0)})$. In this paper, we would restrict ourselves to corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$.
%

Using this parameterized waveform model, which we will call $\pSEOB$ in this paper, we infer bounds on our beyond-GR parameters for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}. We detail these results in Sec.~\ref{sec:results}.
%
%In this paper, we will restrict ourselves to the $k=(2,2,0)$ QNM and corrections to the $j=0$ and $j=1$ terms, i.e., leading and next-to-leading order corrections to the spin, $\chi_f$. Assuming a prior which is uniform on the beyond-GR parameters, $(\ell, \{\delta \omega_k^{(j)}\},\{\delta \tau_k^{(j)}\})$, we employ a Bayesian framework to stochastically sample over the parameter space using the \texttt{LALInference} algorithm.~\cite{lallsuite}.
%
%The bounds on these beyond-GR parameters, for specific cases of modified gravity theories outlined in Sec.~\ref{sec:review_theories}, are detailed in Sec.~\ref{sec:results}.

@@

\subsection{From theory-agnostic to theory-specific QNM results}
\label{sec:theory_specific_qnm}

@@

Now as we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating BHs
is modified gravity theories come in two families depending on their transformation upon a parity
transformation: axial and polar. Which is the one we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work on a theory-by-theory case and in each theory perform a
a translation between the metric perturbations $h_{\mu\nu}$ in the Regger-Wheeler-Zerilli gauge and
connect it with the transverse-traceless gauge used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion in Ref.~\cite{Hui:2021cpm} for a counterexample).
and thus how axial and polar gravitational QNMs appear in the GW signal has to be answered in theory-by-theory.
%
This is outside the scope of this paper and here we take the more pragmatic approach
of simply choosing the \emph{least damped gravitational mode between the two parities.}
%
The justification is this is the mode which, if excited during the merger, is
the most likely to appear in the signal.

@@

%%%%%%%%%%%%%%%%%
\section{Results using LIGO-Virgo events}
\label{sec:results}
%%%%%%%%%%%%%%%%%

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately described by our waveform model \pSEOB{}, we can infer the parameters of the model, $\bm{\lambda}$, given the hypotheis $\mathcal{H}$, using the Bayes theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{P(\bm{\lambda} \vert \mathcal{H}) P(d \vert \bm{\lambda},\mathcal{H})}{P(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$P(\bm{\lambda} \vert \mathcal{H})$ the prior,
$P(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$P(d \vert \mathcal{H})$ the evidence. The set of parameters, $\bm{\lambda}$, detailed in Sec.~\ref{sec:review_pSEOB}
include the set of beyond-GR parameters, $\{\ell, \delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$.
%
As discussed in Secs.~\ref{sec:review_theories} and \ref{sec:theory_specific_qnm}, there is a single coupling
constants in the cases of EdGB and dCS theories, but two in the cubic and
quartic EFTofGR. We also hold the other beyond-GR parameters, $\{\delta \omega_k^{(j)}\},$ and $\{\delta \tau_k^{(j)}\}$
fixed to theory-specific predictions. Hence, our set of beyond--GR parameters only includes the parameter $\{\ell \}$.
%

Assuming Gaussian and stationary noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln \mathcal{L}(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk
denotes complex conjugation and $S_{n}(f)$ is the power spectrum density
of the detector. Assuming a speicifc prior distribution for our parameters, 
we stochastically sample over the parameter space 
using a \texttt{LALInference} Markhov Chain Monte Carlo 
algorithm.~\cite{lallsuite}. We subsequently marginalise over the 
remaining parameters to obtain the posterior probability distribution on $\ell$. 
An example of this is shown in Fig. XX.

\subsection{Event selection}

@@

%We will consider the events GW150915~\cite{LIGOScientific:2016aoc} and GW200129~\cite{LIGOScientific:2021djp}.
%
%They are suitable for ringdown studies due to the high-masses and high signal-to-noise ratios (SNR).
Although the LIGO-Virgo detectors have observed 90 GW signals, we select, for our analyses in this paper, the ones that are likely to be the most promising probes of the BH ringdown using our method. In~\cite{LIGOScientific:2021sio}, the \textit{pSEOBNRv4HM} analysis in Sec.VIII A.2 (which is most similar to the \pSEOB{} model presented in this paper) identified two events which provided the strongest bounds on the dominant QNM measuments -- GW150914~\cite{LIGOScientific:2016aoc} and GW200129~\cite{LIGOScientific:2021djp}.

\ag{event selection, noise, PE settings, priors}
